### 2018-04-02 vipassana-for-hackers.tex
v1.4
added bridge/wall imagery

\begin{titlepage}
   \vspace*{\stretch{1.0}}
   \begin{center}
     \Huge\textbf{Vipassana for Hackers}\\
     \Large\textbf{Instigating the Curious}\\
     \vspace{5cm}
     \large\textit{Steven Deobald}\\
     \large\textit{Version 1.3}\\
     \large\textit\today
   \end{center}
   \vspace*{\stretch{2.0}}
\end{titlepage}

@@

@@

@@

>>>>

\begin{titlepage}
   \vspace*{\stretch{1.0}}
   \begin{center}
     \Huge\textbf{Vipassana for Hackers}\\
     \vspace{5cm}
     \large\textit{Steven Deobald}\\
     \large\textit{Version 1.4}\\
     \large\textit\today
   \end{center}
   \vspace*{\stretch{2.0}}
\end{titlepage}

@@

\begin{center}
  \LARGE{Anapana: The Bridge}
\end{center}

At the end of the first day of a 10-day Vipassana course, the breath is described as a bridge ``from the known to the unknown, because respiration is one function of the body that can be either conscious or unconscious, intentional or automatic.'' During the 3-day Anapana session of a 10-day course, students are instructed to \textit{consciously} direct attention toward their breath -- and even gently control the breath for brief periods to maintain attention when the mind won't stop wandering -- until awareness of breath, without force, lasts for extended periods. Given that the breath is the one function of the body which operates both consciously and unconsciously, it makes sense that it acts as the starting point for many meditation practies, including Vipassana. The intention of Vipassana is to train the mind out of habits and tendencies which have embedded themselves into the brain and nervous system at an unconscious level. To do this, one must first learn to access one's unconscious. Through the breath, Anapana meditation fulfills this purpose.

Vipassana requires a 10-day course to learn. Anapana does not. Before his death, S.N. Goenka released both 70-minute audio instructions as MP3s for the Anapana meditation technique.\cite{anapana} The 70-minute instructions can be found at \url{http://www.vipassana.co/What-is-Anapana} under ``Anapana Meditation for All'' and come as four 15-minute sessions plus a 10-minute ``daily practice'' session.

\begin{figure}[H]
  \centering
  \includegraphics[width=8cm]{images/bridge.png}
  \label{fig:bridge}
\end{figure}

@@

\begin{center}
  \LARGE{Vipassana: The Sledgehammer}
\end{center}

During the Day 4 discourse of a 10-day course, the imagery of the bridge is extended by what is on the far side of that bridge: a giant wall. The wall represents the remaining barrier between the conscious and unconscious mind -- and Vipassana is the tool used to break this barrier. By systematically examining the nervous system, outside-in, one digs through all the layers of unconscious habit as it manifests in bodily sensation.

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.48\linewidth}
    \includegraphics[width=\linewidth]{images/wall.jpg}
  \end{subfigure}
  \begin{subfigure}[b]{0.48\linewidth}
    \includegraphics[width=\linewidth]{images/bridge.png}
  \end{subfigure}
  \label{fig:wall}
\end{figure}

And what comes of this examination? Sensations in the body are exploded into view and what we normally experience as a narrow band of sensation from the roughly unpleasant to the roughly pleasant: intense sensations like abdominal pain and annoying sensations like uncontrollable itching are gross sensations we normally consider unpleasant; intense sensations like sexual contact and light sensations like touching fur are gross sensations we normally consider pleasant.

This explosion of sensation occurs on an entirely new axis. The range can still be considered pleasant to unpleasant but ``pleasant'' begins to cover new, previously unexperienced territory. These experiences have little or no parallel in mundane life and are best described as ``subtle'': fine oscillations and vibration. This new axis of experience is the sledgehammer used to break through the wall separating conscious from unconscious. Unpleasant sensations and pleasant sensations, both, undergo change. As one observes this change, the change itself comes faster and faster and becomes easier and easier to observe. The goal is to avoid reacting to unpleasant sensations with aversion and to avoid reacting to these newfound pleasant sensations with craving. The task is harder than it sounds. However, the calmer a meditator remains while exploring these new depths, the further she explores.

Unfortunately, unlike Anapana meditation, there is no way to learn Vipassana from an MP3. This practice of observing subtler and subtler sensation requires quite a bit of time, assistance from a live teacher, and repeated instruction. The only way one can learn the technique of Vipassana is to take a 10-day course.

@@

\bibitem{anapana}
  S.N. Goenka
  \url{http://www.vipassana.co/What-is-Anapana}
  \textit{What is Anapana?}
  Vipassana Research Institute, 21 October 2016.
